---
title: "Genthat CRAN Experiment"
author: "Filip Krikava & Jan Vitek"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float: true
    highlight: tango
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(stringr)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE, comment=NA)

# genthat options
options(genthat.source_paths="/CRAN")
options(genthat.debug=FALSE)
options(genthat.max_trace_size=getOption("genthat.max_trace_size", 512*1024))

output_dir <- "experiment"
```

```{r aux functions, include=FALSE}
load_coverage <- function(path) {
  coverage <- readRDS(path)
  
  if (is.data.frame(coverage)) {
    # only keep coverage of R file - genthat does not support C/C++ coverage
    coverage_df <- filter(coverage, grepl("(.R|.r)$", filename))
    # compute the percentage
    tests_coverage <- (sum(coverage_df$value > 0) / length(coverage_df$value)) * 100
    data_frame(tests_coverage, tests_coverage_time=as.numeric(attr(coverage, "time"), unit="secs"))
  } else {
    data_frame(tests_coverage=NA, tests_coverage_time=NA)
  }
}

load_genthat_stats <- function(path) {
  tests <- readRDS(path)
  stats <- attr(tests, "stats")
  
  if (!is.null(stats)) {
    data_frame(
      all=stats[1], 
      generated=stats[2], 
      ran=stats[3], 
      kept=stats[4], 
      coverage=stats[5], 
      genthat_tests_time=stats[6], 
      trace_time=as.numeric(attr(tests, "time"), unit="secs")
    )
  } else {
    data_frame(all=NA, generated=NA, ran=NA, kept=NA, coverage=NA, genthat_tests_time=NA, trace_time=NA)
  }
}

fmt <- function(x) format(round(x, 2), big.mark=",")
my_avg <- function(xs) str_c(fmt(mean(xs, na.rm=TRUE)), " (s=", fmt(sd(xs, na.rm=TRUE)), ", m=", fmt(median(xs, na.rm=TRUE)), ")")
```

```{r constants from the paper, include=FALSE}
PAPER_NPACKAGES <- 1547
PAPER_SLOC <- 1700000
PAPER_TEST_COVERAGE_SLOC <- 267113
PAPER_COVERAGE_SLOC <- 704450
PAPER_ACCURACY <- 0.8
```

This notebook processes the results from the CRAN experiment and reproduces the data from Section 4.

## Loading the results

First we load the data from packages that ran.

```{r}
packages_df <- data_frame(
  package_path=list.dirs(output_dir, full.names=TRUE, recursive=FALSE),
  package=basename(package_path),
  coverage_file=file.path(package_path, "coverage.RDS"),
  tests_file=file.path(package_path, "tests.RDS")
) %>% 
  select(-package_path) %>%
  filter(file.exists(coverage_file), file.exists(tests_file))
```

```{r test coverage}
packages_tests_df <- 
  rowwise(packages_df) %>%
  do(load_coverage(.$coverage_file))
```

```{r genthat test extraction}
packages_gt_df <- 
  rowwise(packages_df) %>%
  do(load_genthat_stats(.$tests_file))
```

```{r}
packages_all_df <-
  select(packages_df, package) %>%
  bind_cols(packages_tests_df) %>%
  bind_cols(packages_gt_df)
```

Next, we load the code size.
We ran the [cloc](http://cloc.sourceforge.net/) tool in the docker image to compute the size of R source code in all the packages in `/CRAN` directory.
Each package has `sloc.csv` file in its root which a result of running:

```sh
cloc --quiet --csv %p/R | sed 's/,\".*$//' | tail -n +2 > %p/sloc.csv
```

where `%s` is the name of the project.
Since this is a static information that does nto depend in anyway on our tool, all the files have been already generated during the image build.

```{r}
sloc <- map_dfr(packages_all_df$package, function(x) {
  # read the output of cloc in CSV, skip all but code column
  s <- read_csv(file.path("/CRAN", x, "sloc.csv"), col_types=cols("code"="i"))
  # sum the code per package
  data_frame(package=x, code=sum(s$code))
})
```

## Results

```{r, include=FALSE}
test_coverage_sloc <- sum(sloc$code) * (mean(packages_all_df$tests_coverage, na.rm=T)/100)
coverage_sloc <- sum(sloc$code) * (mean(packages_all_df$coverage, na.rm=T)/100)
accuracy <- sum(packages_all_df$ran)/sum(packages_all_df$all)
```

Together, this experiment ran `r length(readLines("packages.txt"))` packages (the number of lines in `packages.txt`) out of which `r nrow(packages_all_df)` packages completed.
This is `r fmt(nrow(packages_all_df)/PAPER_NPACKAGES*100)`% of the packages reported in the the paper.
These packages comprises of `r fmt(sum(sloc$code))` lines of code (excluding whitespace and comments), which is `r fmt(sum(sloc$code)/PAPER_SLOC*100)`% of the original experiment.
We selected the packages which had the original tracing time less than 2 minutes to keep the artifcat time manageable.

### Summary

The average test code coverage is `r my_avg(packages_all_df$tests_coverage/100)`.
With genthat this is increased to `r my_avg(packages_all_df$coverage/100)`.
In terms of source lines of code (as reported in both the paper abstract and paper introduction), **genthat improved code coverage from the original rather low value of `r fmt(test_coverage_sloc)` lines to `r fmt(coverage_sloc)` lines**.
This is an increase of `r fmt(coverage_sloc/test_coverage_sloc*100)`%.
In the paper we reported `r fmt(PAPER_COVERAGE_SLOC/PAPER_TEST_COVERAGE_SLOC*100)`%.

The major results in the paper are in Table 1, Table 2 and Figure 4 which are shown bellow.

Table 3 describes what are the cases of lost accurracy.
A part of the table requires manual inspection of the failures (how to get errors has been explained in the _overview_ notebook).
Therefore, we reports only the overall accuracy `r fmt(accuracy*100)`% which should be similar to what we reported in the paper `r fmt(PAPER_ACCURACY*100)`%.
This of course depends on the package selection, yet in a resonable number of packages the numbers shall be close.

Finally, Figure 5 with reverse dependencies is skipped because:

1. rerunning reverse dependencies for these packages will be highly time consuming,
2. we have already shown in the _overview_ how it is done on a single package `Rvmmin`,
3. the above experiment shows the strategy of running genthat in a batch.

### Table 1

```{r}
table1 <- tribble(
  ~Name,                      ~Overall,                            ~`Average per package`,
  "Traced unique calls",      fmt(sum(packages_all_df$all)),       my_avg(packages_all_df$all),
  "Generated tests",          fmt(sum(packages_all_df$generated)), my_avg(packages_all_df$generated),
  "Valid and Correct tests",  fmt(sum(packages_all_df$ran)),       my_avg(packages_all_df$ran),
  "Non-redundant tests",      fmt(sum(packages_all_df$kept)),      my_avg(packages_all_df$kept),
  "Ratio of reproduced test", fmt(accuracy),                       my_avg(packages_all_df$ran/packages_all_df$all)
)

table1
```

### Figure 4

```{r}
select(packages_all_df, package, coverage, tests_coverage) %>%
  mutate(Default=tests_coverage/100, Genthat=coverage/100) %>%
  gather(key="key", value="value", Default, Genthat) %>%
  ggplot(aes(x=value, fill=factor(key, levels=c("Genthat", "Default")))) + 
    geom_histogram(alpha=.4, bins=100, colour="grey") + 
    scale_x_continuous(labels=scales::percent_format()) +
    theme_minimal() +
    theme(legend.position=c(.9,.75)) +
    labs(x="Code coverage", y="Number of packages", fill="Tests")
    
```

### Table 2

```{r}
table2 <- tribble(
  ~Name,                  ~`Average per package`,
  "Tracing",              my_avg(packages_all_df$trace_time),
  "Testing with genthat", my_avg(packages_all_df$genthat_tests_time),
  "Testing",              my_avg(packages_all_df$tests_coverage_time)
)

table2
```
