---
title: "Genthat Overview"
author: "Filip Krikava & Jan Vitek"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float: true
    highlight: tango
---

```{r setup, include=FALSE}
# this is required and Rscript will not load
library(methods)

knitr::opts_chunk$set(echo = TRUE, comment=NA, cache=TRUE)

# remove previous artefacts
if (dir.exists("tmp")) {
  unlink("tmp", recursive=TRUE)
}

# install local packages
install.packages(
  list.dirs("packages", recursive=FALSE, full.names=TRUE),
  repos=NULL, 
  INSTALL_opts = c("--example", "--install-tests", "--with-keep.source", "--no-multiarch"),
  quiet=T
)

fmt <- function(x) format(round(x, 2), big.mark=",")
```

In this R notebook, we provide an overview of the genthat package.
We start with its low-level API to show how a test is extracted from a function call using examples based on the code shown in sections 2 and 3 in the paper.
Next, we show how the test extraction process can be automated and run on a whole package including its reverse dependencies.

## Basics

### Setup

We start by defining a few functions to setup our playground.
A `filter` function and a few predicates:

```{r}
filter <- function(xs, p) xs[sapply(xs, p)]

is_even <- function(x) x %% 2L == 0L
is_odd <- function(x) !is_even(x)
```

Given a sequence of integers:

```{r}
nums <- 1L:10L
nums
```

we can now get all the even:

```{r}
filter(nums, is_even)
```

as well as all the odd numbers:

```{r}
filter(nums, is_odd)
```

### Loading genthat

In order to use genthat, we need to load it.
It is a regular R package that has been already installed in the docker image so we directly use the `library` function:

```{r}
library(genthat)
```

### Function tracing

Genthat extracts test from _traces_, which are recorded function invocations.
To capture invocations of a function, we first need to decorate it. 
This is done by the `decorate_function` that modifies the original function body to additionally records its arguments, return values, and the current state of the random number generator.
Before calling the function we can double check the body of the `filter` function defined above:

```{r}
body(filter)
```

As expected, it is exactly what we have written at the beginning.
After calling:

```{r}
decorate_function(filter)
```

the body is changed to:

```{r}
body(filter)
```

The inserted code has two parts: a prologue that runs before any of the original function’s code and an epilogue that runs right before the function exits. 
The latter relies on the `on.exit` hook provided by R.
The prologue records the random seed.
Due to R’s lazy nature, recording of the argument values is done at the end to avoid forcing a promise before it is needed—that might cause a side effect and change the result of the program or cause an error.
With the exception of environments, R values are immutable with a copy on write semantics, so any modification of function arguments in the function’s body will trigger their duplication.
There are two things that need to be done before recording call details.
First, tracing must be temporally disabled.
Next, we need to check whether the function terminated normally or threw an exception.
A call is only recorded in the case the function did not fail.

From now, when the function is called and terminates normally, its invocation is recorded in a trace.

```{r}
filter(nums, is_odd)
```

To see the recorded traces, we use `copy_traces()` which returns a list of the traces seen so far (since the last reset, _cf._ bellow).
Internally, genthat uses a hash map, storing only unique traces (unique arguments and return value).

There is just one trace:

```{r}
str(copy_traces())
```

A trace has 6 fields:

- `fun` - the called function name
- `pkg` - the name of the package the `fun` comes from (`NULL` for global definitions)
- `args` - values of arguments
- `globals` - values of free variables
- `seed` - the random seed
- `retv` - teh return value

We can see that the `is_odd` predicate is captured as a symbol in the arguments and then its actual value is stored in the list of global variables.
In the end we want the call to be fully reproducible in a standalone test that can run in a clean R session.
For this, genthat has to capture all the free variables in a call expression.
In this case, it captured the `is_odd` predicate, but recall, that `is_odd` was defined in terms of `is_even`:

```{r}
body(is_odd)
```

Therefore, the trace related to the call `filter(nums, is_odd)` should also keep the definition of the `is_even` function.
Since the `is_even` is only related to the `is_odd` function, not to the `filter` function, it is stored in the function environment.

We can list all the symbols that are in the environment of the `is_odd` function that is embedded in the trace:

```{r}
ls.str(environment(copy_traces()[[1]]$globals$is_odd))
```

as well the code:

```{r}
body(get("is_even", environment(copy_traces()[[1]]$globals$is_odd)))
```

which shall be the same as in the `is_even` defined at the beginning.

### Test generation

Now, when we have traces, we can turn them into tests.
This is done using by `generate_test` method that returns a string vector with the extracted test.

```{r}
test <- generate_test(copy_traces()[[1]])
cat(paste(test, collapse="\n\n"))
```

The `.ext.seed` represents the state of the random number generator before the call.
This variable is part of what genthat defines as _external_ variables, _i.e._ variables that are not defined in the source file, but rather stored in an external file in RDS format (R binary serialization).
The reason is that some variables can be rather large and having them directly in a source file
would hamper its readability.

The rest of the test is straight forward.
It defines all the free variables from the call expression as well as the call to `expect_equal` asserting the actual and expected value.

#### External values

The string returned from `generate_test` includes an attribute `externals` that is an environment contains all external variables referenced by the test.

```{r}
ls.str(attr(test, "externals"), all.names=TRUE)
```

### Running generated test

The generate tests are in the [testthat](https://github.com/r-lib/testthat) format, a popular unit testing library for R.
Because of the external variables, the tests need to be run in a correct environment.
Genthat provides `generate_test_file` which stores generated tests and external variables in files.

```{r}
# save generated test in a file
output_dir <- tempfile()
test_file <- generate_test_file(copy_traces()[[1]], output_dir)
```

The structure is `<package-name>/<function-name>/test-<seq>.{R,ext}` where `seq` is just an increasing number sequence. The `.R` file contains the test code and the `.ext` file stores the external variables.

In our example, since we have no package name (the `filter` is defined in a global environment), the package name is `_NULL_` to emphasize this fact.

```{r}
cat(paste(system2("tree", output_dir, stdout=TRUE), collapse="\n"))
```

The `generated_test_file` returns the file path. This can be fed into `test_generated_file`, which is a thin wrapper over `testthat::test_file` function additionally taking care of the external variables.

```{r}
result <- test_generated_file(test_file)
result
```

There is also a `run_generated_tests` function that takes a vector of test files to run consecutively. It returns vector of elapsed time as well as any encountered errors. This is useful for batch processing.

```{r}
run_generated_test(test_file)
```

### Cleaning up

Genthat provides two functions to clean up:

- `reset_traces` to remove all recorded traces:
```{r}
reset_traces()
str(copy_traces())
```

- `reset_function` to restore the original function body:
```{r}
reset_function(filter)
body(filter)
```

## Extracting tests from a package

Genthat is usually used in a batch, extracting tests for a whole package.
To show how this work, we will use [Rvmmin](https://cran.r-project.org/web/packages/Rvmmin/index.html), an existing R package for variable metric nonlinear function minimization with bounds constraints.
It has over 1000 lines of code and has been on CRAN for a couple of years.
The package source is in `packages/Rvmmin`:

```{r}
cat(paste(system2("tree", "packages/Rvmmin", stdout=TRUE), collapse="\n"))
```

We can see that the package defines a test (`tests/BTbad.R`), it has some documentation (`man/*.Rd`) and a vignette (`vignettes/Rvmmin.Rmd`).

First, we checking the test code coverage of the package using [covr](https://cran.r-project.org/web/packages/covr/):
```{r}
library(covr)
tests_coverage <- package_coverage("packages/Rvmmin", type="tests")
tests_coverage_value <- compute_coverage(tally_coverage(tests_coverage))
tests_coverage_value
```
From the coverage of `r fmt(tests_coverage_value)`% we see that the tests do not cover much in this package.

Since there seems to be some runnable code (examples in documentation and vignettes), we can try to improve this using genthat.
First, we define where genthat finds the source code of packages:

```{r set source code location}
options(genthat.source_paths="packages")
```

Running the test extraction is done using `gen_from_package` which basically orchestrates the different functions we have showed in the previous section.

```{r tracing Rvmmin package}
result <- 
  gen_from_package(
    "Rvmmin", 
    types="all", 
    action="generate", 
    prune_tests=TRUE, 
    output_dir="tmp"
  )
```

- `types` describes which code to to run (`vignettes`, `tests`, `examples`)
- `action` - `generate` - generated testthat unit tests, `export` - stores raw traces
- `prune_test` - discards the tests that do not increase code coverage
- `output_dir` - where to generate the tests

It generated `r nrow(result)` tests, with the code coverage of `r fmt(max(result$coverage))`%.

Example 
```{r}
result
```

- `file` - where the code generating this test originates
- `output` - path to the test
- `elapsed` - test running time
- `coverage` - test coverage computed as a sum of previous tests plus this test

The reported test coverage is only for the tests generated by genthat.
Since, genthat also runs package test code, it should also capture all their calls.
However, in some cases genthat cannot properly trace a call and so it can be useful to include also the original tests in the final suite.
The `compute_coverage` function can compute test coverage from multiple sources.
To combine the original test coverage as well as the coverage from genthat tests, we need the raw coverage indicating which files, functions and line numbers we hit during an evaluation.
This is included in the `raw_coverage` attribute in the result from `gen_from_package`.

```{r}
coverage <- compute_coverage(tally_coverage(tests_coverage), attr(result, "raw_coverage"))
coverage
```

In this case, the code coverage is the same meaning that genthat managed to capture all the calls from the package's tests.
This is however not always the case as we reproted in the paper in Table 1 and Table 3.

The tests were generated in `tmp/Rvmmin`:

```{r}
cat(paste(system2("tree", "tmp/Rvmmin", stdout=TRUE), collapse="\n"))
```

Next to `raw_coverage`, the `result` return value also contains some additional attributes:

`stats` provides overview of the test extraction:

```{r}
attr(result, "stats")
```

where

- `all` is the number of unique traces (unique arguments / return values)
- `generated` is the number of successfully generated tests
- `ran` is the number of tests that passed
- `kept` is the number of tests that were kept after the pruning based on coverage
- `coverage` is the resulting coverage of R code in %
- `elapsed` is the time it took to run the generated tests

The `errors` provides details about the failures in tracing, tests generation and test running:

```{r}
head(attr(result, "errors"))
```

which is is useful for debugging.
This was the source for writing Table 3 in the paper.

## Extracting tests from reverse dependencies

Genthat can also be use to extract tests for a package from code that comes from the package's 
reverse dependencies. In this case we already have pretty high code coverage, but we can try to
boost it up a bit more.

First, we need to find what are the reverse dependencies:

```{r}
rdeps <- tools::package_dependencies("Rvmmin", which=c("Depends", "Imports", "LinkingTo", "Suggests"), reverse=TRUE, recursive=FALSE)
rdeps <- unlist(rdeps, use.names=FALSE)
rdeps
```

We use only direct reverse dependencies. While we can use also indirect ones, it usually does not bring much and only takes a lot of time to run.

The call to `gen_from_package` is almost the same as with only the package code with the
difference that we specify what packages to run:

```{r tracing reverse dependencies}
result_rdeps <- 
  gen_from_package(
    pkgs_to_trace="Rvmmin",
    pkgs_to_run=c("Rvmmin", rdeps),
    types="all", action="generate", prune_tests=TRUE, output_dir="tmp")
```

Looking at

```{r}
coverage_rdeps <- compute_coverage(attr(result_rdeps, "raw_coverage"))
attr(result_rdeps, "stats")
```

we can see that the running the reverse dependencies add a `r nrow(result_rdeps) - nrow(result)` more tests and 
increased the coverage from `r fmt(max(coverage))`% to `r fmt(coverage_rdeps)`%.

## Appendices

Following are two appendices that replicate the additional examples shown in
the Section 3 of the paper.

### Example from Section 3.1

```{r}
decorate_function(filter)
```

The example is the following:

```{r}
m <- 0.5
n <- 1
filter(runif(10) + m, function(x) x > n)
```

It produces a familiarly looking trace:

```{r}
str(copy_traces())
```

```{r}
reset_traces()
```

If we change the predicate to be a standalone function:

```{r}
gt <- function(x) x > n
filter(runif(10) + m, gt)
```

the trace will have also include the `gt` definition

```{r}
str(copy_traces())
```

with the `n` value in its environment:

```{r}
ls.str(environment(copy_traces()[[1]]$globals$gt))
```

### Example from Section 3

Before we continue with one more example, we just remove all the traces, 
to have a fresh tracing environment:

```{r}
reset_traces()
str(copy_traces())
```

In Figure 2, in Section 3, there is the following code:

Code from package `L`:

```{r}
f <-function(x, y, l=c(deparse(substitute(x)), deparse(substitute(y)))) {
  df <- data.frame(dx=x, dy=y)
  attr(df,"label") <- l
  df
}
```

Decorate function `f`:

```{r}
decorate_function(f)
reset_traces()
```

Client code:

```{r results='hide'}
a <- c(4, 4)
b <- 1:2
f(sqrt(a) + 1, b)
```

Resulting tests:

```{r}
test <- generate_test(copy_traces()[[1]])
cat(paste(test, collapse="\n\n"))
```