---
title: "Genthat Overview"
author: "Filip Krikava & Jan Vitek"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float: true
    highlight: tango
---

```{r setup, include=FALSE}
# this is required and Rscript will not load
library(methods)

knitr::opts_chunk$set(echo = TRUE, comment=NA)

# remove previous artefacts
if (dir.exists("tmp")) {
  unlink("tmp", recursive=TRUE)
}

# install local packages
install.packages(
  list.dirs("packages", recursive=FALSE, full.names=TRUE),
  repos=NULL, 
  INSTALL_opts = c("--example", "--install-tests", "--with-keep.source", "--no-multiarch"),
  quiet=T
)

fmt <- function(x) format(round(x, 2), big.mark=",")
```

In this R notebook, we provide an overview of genthat package. We start with its 
low-level API to show how a test is extracted from a function call using examples 
based on the code shown in sections 2 and 3 in the paper. Next, we show how the 
test extraction process can be automated and run on a a whole package.

## Basics

### Setup

We start by defining a few functions to setup our playground.
A `filter` function and a few predicates:

```{r}
filter <- function(xs, p) xs[sapply(xs, p)]

is_even <- function(x) x %% 2L == 0L
is_odd <- function(x) !is_even(x)
```

Given a sequence of integers:

```{r}
nums <- 1L:10L
nums
```

we can now get all the even:

```{r}
filter(nums, is_even)
```

as well as all the odd numbers:

```{r}
filter(nums, is_odd)
```

### Loading genthat

In order to use genthat, we need to load it.
It is a regular R package that has been already installed in the 
docker image so we directly use the `library` function:

```{r}
library(genthat)
```


### Function tracing

Genthat extracts test from _traces_, which are recorded function invocations.
To capture invocations of a function, we need to decorate it. 
The `decorate_function` modifies the original function to additionally records
its arguments, return values and the current state of the random number
generator.

Decorating is done using `decorate_function`. Before calling that we can 
double check the body of the `filter` function defined above:

```{r}
body(filter)
```

As expected, it is exactly what we have written at the beginning.
After calling:

```{r}
decorate_function(filter)
```

the body will be changed:

```{r}
body(filter)
```

The inserted code has two parts: a prologue that runs before any
of the original function’s code and an epilogue that runs right
before the function exits. The latter relies on the on.exit hook
provided by R. The prologue records the random seed. Due to R’s
lazy nature, recording of the argument values is done at the end
to avoid forcing a promise before it is needed—that might cause
a side effect and change the result of the program or cause an
error. With the exception of environments, R values are immutable
with a copy on write semantics, so any modification of function
arguments in the function’s body will trigger their duplication.
There are two things that need to be done before recording call
details. First, tracing must be temporally disabled. Next, we need
to check whether the function terminated normally or threw an
exception. A call is only recorded in the case the function did not
fail.

From now, when the function is called and terminates normally,
its invocation is recorded in a trace.

```{r}
filter(nums, is_odd)
```

To see the recorded traces, we can use `copy_traces()` which returns a list
of the traces seen so far. Internally, genthat uses a hash map, storing only
unique traces.

There shall be just one trace:

```{r}
str(copy_traces())
```

A trace has 6 fields:

- `fun` - the called function name
- `pkg` - the name of the package the `fun` comes from (`NULL` for global definitions)
- `args` - values of arguments
- `globals` - values of free variables
- `seed` - random seed
- `retv` - return value

We can see that the predicate is captured as a symbol
in the arguments and then its actual value is stored in the list of global 
variables. 

Recall, that the `is_odd` predicate was defined in terms of `is_even`:

```{r}
body(is_odd)
```

therefore, the trace related to the call `filter(nums, is_odd)` should also keep
the definition of the `is_even` function. Since the `is_even` is only related to
the `is_odd` function, not to the `filter` function, it is stored in the function
environment.

We can list all the symbols that are in the environment of the 
`is_odd` function that is embedded in the trace.
```{r}
ls.str(environment(copy_traces()[[1]]$globals$is_odd))
```

As well the code:

```{r}
get("is_even", environment(copy_traces()[[1]]$globals$is_odd))
```

which shall be the same as in the `is_even` defined at the beginning 
(except the extra marking attribute used by genthat).

### Test generation

Now, when we have traces, we can turn them into tests.
This is done using by `generate_test` method which returns a string vector with
the extracted test.

```{r}
test <- generate_test(copy_traces()[[1]])
cat(paste(test, collapse="\n\n"))
```

The `.ext.seed` represents the state of the random number generator before the
call. This variable is part of what genthat defines as _external_ variables, _i.e._
variables that are not defined in the source file, but rather stored in an 
external file in RDS format (R binary serialization). The reason is that 
some variables can be rather large and having them directly in a source file
would hamper its readability.

The string returned from `generate_test` includes an attribute `externals` that
is an environment contains all external variables referenced by the test.

```{r}
ls.str(attr(test, "externals"), all.names=TRUE)
```

### Running generated test

The generate tests are in the [testthat](https://github.com/r-lib/testthat) format, 
a popular unit testing library for R. Because of the external variables, the tests
need to be run in a correct environment. Genthat provides `generate_test_file`
which stores generated tests and external variables in files.

```{r}
# save generated test in a file
output_dir <- tempfile()
test_file <- generate_test_file(copy_traces()[[1]], output_dir)
```

The structure is `<package-name>/<function-name>/test-<seq>.{R,ext}` where
`seq` is just an increasing number sequence. The `.R` file contains the test code and
the `.ext` file stores the external variables.

In our example, since we have no package name (the `filter` is defined in a global environment),
the package name is `_NULL_` to emphasize this fact.

```{r}
cat(paste(system2("tree", output_dir, stdout=TRUE), collapse="\n"))
```

The `generated_test_file` returns the file path. This can be fed into
`test_generated_file`, which is a thin wrapper over `testthat::test_file`
function taking care of the external variables.

```{r}
result <- test_generated_file(test_file)
result
```

There is also a `run_generated_tests` function that takes a vector of test files
to run consecutively. It returns vector of elapsed time as well as
any encountered errors. This is useful for batch processing.

```{r}
run_generated_test(test_file)
```

### Cleaning up

Genthat provides two functions to clean up:

- `reset_traces` to remove all recorded traces:

```{r}
reset_traces()
str(copy_traces())
```

- `reset_function` to restore the original function body:

```{r}
reset_function(filter)
```


```{r}
body(filter)
```

## Extracting tests from a package

Genthat is usually used in a batch processing, extracting tests for a whole
package. To show how this work, we will use [Rvmmin](https://cran.r-project.org/web/packages/Rvmmin/index.html),
an existing R package for variable metric nonlinear function minimization with bounds constraints.
It has over 1000 lines of code and has been on CRAN for a couple of years.
The package source is in `packages/Rvmmin`:

```{r}
cat(paste(system2("tree", "packages/Rvmmin", stdout=TRUE), collapse="\n"))
```

We can see that the package defines a test (`tests/BTbad.R`), it has some documentation (`man/*.Rd`)
and a vignette (`vignettes/Rvmmin.Rmd`).

Checking the test code coverage:
```{r}
library(covr)
package_coverage("packages/Rvmmin", type="tests")
```
we see that the test does not cover much.

Since there seems to be some runnable code (examples in documentation and vignettes), we can try to improve this using genthat.
First, we define where genthat finds the source code of packages:

```{r}
options(genthat.source_paths="packages")
```

Running the test extraction is done using `gen_from_package` which basically
orchestrates the different functions we have showed in the previous section.

```{r}
result <- 
  gen_from_package("Rvmmin", types="all", action="generate", prune_tests=TRUE, output_dir="tmp")
```

It generated `r nrow(result)` tests, with the code coverage of `r fmt(max(result$coverage))`%.

```{r}
result
```

The tests were generated in `tmp/Rvmmin`:

```{r}
cat(paste(system2("tree", "tmp/Rvmmin", stdout=TRUE), collapse="\n"))
```

The `result` return value also contains some attributes with additional information.
The `stats` provides overview of the test extraction:

```{r}
attr(result, "stats")
```

where

- `all` is the number of unique traces (unique arguments / return values)
- `generated` is the number of successfully generated tests
- `ran` is the number of tests that passed
- `kept` is the number of tests that were kept after the pruning based on coverage
- `coverage` is the resulting coverage of R code in %
- `elapsed` is the time it took to run the generated tests

The `errros` provides details about the failures in tracing, tests generation and test running:

```{r}
head(attr(result, "errors"))
```

This is useful for debugging.

## Extracting tests from reverse dependencies

Genthat can also be use to extract tests for a package from code that comes from the package's 
reverse dependencies. In this case we already have pretty high code coverage, but we can try to
boost it up a bit more.

First, we need to find what are the reverse dependencies:

```{r}
rdeps <- tools::package_dependencies("Rvmmin", which=c("Depends", "Imports", "LinkingTo", "Suggests"), reverse=TRUE, recursive=FALSE)
rdeps <- unlist(rdeps, use.names=FALSE)
rdeps
```

We use only direct reverse dependencies. While we can use also indirect ones, it usually does not bring much and only takes a lot of time to run.

The call to `gen_from_package` is almost the same as with only the package code with the
difference that we specify what packages to run:

```{r}
result_rdeps <- 
  gen_from_package(
    pkgs_to_trace="Rvmmin",
    pkgs_to_run=c("Rvmmin", rdeps),
    types="all", action="generate", prune_tests=TRUE, output_dir="tmp")
```

Looking at

```{r}
attr(result_rdeps, "stats")
```

we can see that the running the reverse dependencies add a `r nrow(result_rdeps) - nrow(result)` more tests and 
increased the coverage from `r fmt(max(result$coverage))`% to `r fmt(max(result_rdeps$coverage))`%.

## Appendices

Following are two appendices that replicate the additional examples shown in
the Section 3 of the paper.

### Example from Section 3.1

```{r}
decorate_function(filter)
```

The example is the following:

```{r}
m <- 0.5
n <- 1
filter(runif(10) + m, function(x) x > n)
```

It produces a familiarly looking trace:

```{r}
str(copy_traces())
```

```{r}
reset_traces()
```

If we change the predicate to be a standalone function:

```{r}
gt <- function(x) x > n
filter(runif(10) + m, gt)
```

the trace will have also include the `gt` definition

```{r}
str(copy_traces())
```

with the `n` value in its environment:

```{r}
ls.str(environment(copy_traces()[[1]]$globals$gt))
```

### Example from Section 3

Before we continue with one more example, we just remove all the traces, 
to have a fresh start:

```{r}
reset_traces()
str(copy_traces())
```

In Figure 2, in Section 3, there is the following code:

Code from package `L`:

```{r}
f <-function(x, y, l=c(substitute(a), substitute(b))) {
  df <- data.frame(dx=x, dy=y)
  attr(df,"label") <- l
  df
}
```

Decorate function `f`:

```{r}
decorate_function(f)
reset_traces()
```

Client code:

```{r results='hide'}
a <- c(4, 4)
b <- 1:2
f(sqrt(a) + 1, b)
```

Resulting tests:

```{r}
test <- generate_test(copy_traces()[[1]])
cat(paste(test, collapse="\n\n"))
```