---
title: "ISSTA Artifact"
authors: "Filip Krikava & Jan Vitek"
output:
  html_document:
    highlight: tango
    keep_md: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NA)
```

```{r constants from the paper, include=FALSE}
PAPER_NPACKAGES <- 1547
```

## Artifact description

This is the artifact for the paper __Tests from Traces: Automated Unit Test Generation for R__ published at ISSTA 2018.
The aim is to:

1. show how the Genthat tool introduced in the paper works and can be used for extracting code from R packages, and
2. reproduce the experiment used to evaluate it.

Concretely, we rerun the all the code snippets shown in Section 3 of the paper and the evaluation from Section 4.

The artifact is composed of two parts:

- Genthat Overview
- CRAN Experiment

The first part corresponds to the Section 3 of the paper where we introduce the tool.
The second part reproduces the experiment that we used to evaluate the tool in Section 4 and reproduce Table 1,2 and Figure 4.

Both parts require the docker image to be run.

## Prerequisites

- git
- [docker](https://www.docker.com/community-edition)
- about 8GB of disk space

## Running the artifact

1. Clone the repository [fikovnik/ISSTA18-artifact](https://github.com/fikovnik/ISSTA18-artifact):
```sh
git clone https://github.com/fikovnik/ISSTA18-artifact
```
The artifact is provided as a docker image `prlprg/issta18-genthat`.
The image contains R 3.4 with a web version of Rstudio and a set of sample packages.

2. Run the docker image:
```sh
./run.sh
```
from the cloned repository.
This runs the docker container in foreground making it easy to close by just issuing CTRL+C or simply closing the terminal window.
It will download the already prebuilt docker image from [docker hub](https://hub.docker.com/r/fikovnik/issta18-artifact/).
If you prefer to build your own, please follow the instructions at the end of this document.
If you ran the docker image manually, please make sure its name is `issta18-genthat`.
The rest of the commands shall be done in new terminal window.

### Genthat Overview

This part essentially explains how genthat works and reproduces all the examples that are in the paper.
There are two ways to run it:

1. Interactively via Rstudio that shall be available on [http://localhost:8787](http://localhost:8787).
   The username is `rstudio` and the password is `rstudio`.
   In the Rstudio application, please open the `overview.Rmd`.
   A [Rmd](https://rmarkdown.rstudio.com/) or RMarkdown file is simply a document in Markdown format that embeds snippets of R code that can be ran (_knitted_ - terminology used in RMarkdown) to produce an output with all the code evaluated.
   This can be done either running by running the snippets _chunk-by-chunk_ (using the green play button next to each snippet) or by running the whole document by pressing Ctrl+Alt+K (Command+Shift+K on OSX). 
   Resulting HTML outputs are already provided in the repository---[`overview.html`](http://localhost:8787/files/overview.html).
   
2. Simply evaluate the whole document by running from the cloned repository:
```sh
docker exec -u rstudio -w /home/rstudio issta18-genthat Rscript -e 'rmarkdown::render("overview.Rmd")'
```
and then open the [`overview.html`](http://localhost:8787/files/overview.html).

### CRAN experiment (Section 4)

The Section 4 presents an experiment that runs genthat on `r PAPER_NPACKAGES` [CRAN](https://cran.r-project.org/) packages.
In this artifact, we present the same one, with the difference that only `r length(readLines("packages-all.txt"))` packages are selected.
The reason is that tracing, as indicated in the paper, takes a long time to run.
These packages represent all the original ones for which tracing took less than 2 minutes.
Running the original package set took **1d 18h** on two virtual nodes with 60GB RAM and 16 CPU at 2.2GHz.
On an average laptop or desktop, it might take much more.

This experiment is composed in a way that gives some flexibility as how much time one wants to devote to the reproduction.
We have created the following package files (package name per line) with different sizes:

- `packages-100.txt` ~ 100 packages, completes in ~30min
- `packages-200.txt` ~ 200 packages, completes in ~45min
- `packages-400.txt` ~ 400 packages, completes in ~75min
- `packages-all.txt` ~ all packages, completes in ~100min

which simply restrict the selection of packages.
The time needed was measured on i7 laptop with 16GB of RAM.
The experiment can also be run iteratively, each time adding more packages.
It does not rerun the package for which it has produced data.
This is why you need to remove the `experiment` directory at the very beginning.

By default the experiment is set to run all packages.
This can be changed this by symlinking the `packages.txt` to a different target set.

It is possible to try genthat on any package.
Simply just add your favorite R package(s) into `packages.txt` file.
Just make sure that they are installed like (basically including their runnable code):

```r
install.packages("MyFavoritePackage", destdir='/CRAN', INSTALL_opts=c('--example', '--install-tests', '--with-keep.source', '--no-multiarch'), dependencies=TRUE)
```

and the downloaded sources in `/CRAN/*.tar.gz` are extracted in `/CRAN`.

#### Rerunning the experiment

The repository already contains the experiment ran (in `experiment` directory). 
To redo the experiment:

1. Make sure the docker image is running

2. Remove the existing results
```sh
rm -fr experiment
```

3. Select the size of the experiment
For example, to run a smaller set
```sh
rm packages.txt
ln -s packages-400.txt packages.txt
```

4. Run
```sh
./run-experiment.sh
```
This runs the `./extract-tests.R` it in [parallel](https://www.gnu.org/software/parallel/) for each package found in the `packages.txt` file.
The result are two RDS files:  
- a `experiment/<package_name>/coverage.RDS` - contains test code coverage from covr  
- a `experiment/<package_name>/tests.RDS` - contains the result of running `gen_from_package`  
plus all the regular genthat extracted tests.  
**This will take a while!**

4. Knit the `cran-experiment.Rmd` using provided [R studio](http://localhost:8787) or run
```sh
docker exec -u rstudio -w /home/rstudio issta18-genthat Rscript -e 'rmarkdown::render("cran-experiment.Rmd")'
```
and then open the resulting [`cran-experiment.html`](http://localhost:8787/files/cran-experiment.html)

#### A note on running package code

Genthat extracts tests by running available R code from R packages.
This is not always deterministic since some of the code tries to access various web pages or have some hard-coded values.
This is means that it might happen that some of the packages might fail.
However, there should be a small number of them.

## Build image locally

The image is hosted on the [docker hub](https://hub.docker.com/r/fikovnik/issta18-artifact/) and shall be available for download.
To make a local build run `build.sh`:

```{sh echo=FALSE}
cat build.sh
```

## Toubleshooting

### Docker on OSX

It is better to use homebrew cask to install docker:

```sh
brew cask install docker
```

in case you see `docker: Cannot connect to the Docker daemon at unix:///var/run/docker.sock. Is the docker daemon running?.` error message
cf: https://stackoverflow.com/a/44719239

### Docker on Linux

In some distribution the package does not add the current user to `docker` group.
In this case, either add yourself to `docker` group or run the `./run*` commands with `sudo`.